A linear inside-outside algorithm for correcting sequencing errors in structured RNA sequences

### 2012-10-07 Recomb/introduction_RECOMB.tex
Quelques typos, révision de méthods

We design a new scoring scheme combining nearest-neighbor models \cite{Turner2010} to isostericity metrics \cite{Stombaugh2009}.
Classical approaches use a Boltzmann distribution whose weights are estimated using a nearest-neighbour energy model~\cite{Turner2010}. However, the
latter only accounts for  canonical and wobble, base pairs. As was shown by Leontis and Westhof~\cite{Leontis2001},
the diversity of base pairs observed in tertiary structures is much larger, albeit their energetic contribution remains unknown. To quantify geometrical differences, 
an isostericity distance has been designed \cite{Stombaugh2009}, increasing as two base pairs geometrically differ from each other in space. Therefore, we 
incorporate these scores in the Boltzmann weights used by \RNApyro.

>>>>

We design a new scoring scheme combining nearest-neighbor models \cite{Turner2010} to isostericity metrics \cite{Stombaugh2009}.
Classical approaches use a Boltzmann distribution whose weights are estimated using a nearest-neighbour energy model~\cite{Turner2010}. However, the
latter only accounts for  canonical and wobble, base pairs. As was shown by Leontis and Westhof~\cite{Leontis2001},
the diversity of base pairs observed in tertiary structures is much larger, albeit their energetic contribution remains unknown. To quantify geometrical discrepancies, 
an isostericity distance has been designed \cite{Stombaugh2009}, increasing as two base pairs geometrically differ from each other in space. Therefore, we 
incorporate these scores in the Boltzmann weights used by \RNApyro.

### 2012-10-07 Recomb/methods_RECOMB.tex
Quelques typos, révision de méthods

\subsection{Probabilistic model}
Our probabilistic model aims at capturing both the stability of the folded RNA and its ability to adopt a predefined 3D conformation.
To that purpose, a Boltzmann weighted distribution is used, based on a pseudo-energy function $\PE{\cdot}$ which includes contributions for both 
the free-energy and the isostericity. 
More precisely, let $\Omega$ be an gapless RNA alignment, $S$ its associated secondary structure, 
then each sequence $s$ has probability proportional to its Boltzmann factor
\begin{align*}
  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT} &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
\end{align*}
where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ are the free-energy and isostericity contributiuons, further described below, and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is the tabulated free-energy of a stacking pair in the Turner model if available, $0$ if $ab=\varnothing$ (no base-pair to stack onto), and using default value $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$). This latter parameter allows to choose whether to completely forbid a sequence inducing invalid base pairs ($\beta = \infty$), or to simply penalize it.
The imprecision introduced by this simplification of the Turner model remains reasonable, since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsection{isostericity contribution}
The concept of isostericity captures the geometric superimposability of two base-pairs, using individual scores given by Stombaugh~\emph{et al}~\cite{Stombaugh2009} based on a probabilistic model. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gapless RNA alignment $\Omega$,  the contribution of the isostericity to the pseudo-energy is given by
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence in comparison with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the alignment sequences.

\subsection{Mutational profiles of sequences}

Let $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the 
{SCFG} generating  
all  secondary structures of length  $|s|$. We are interested in  
the probability of a given position being a specific nucleotide,
over all sequences having at most $m$ mutations from $s$, under the SCFG derivation $S$ 
(formally 
$\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define as a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former presented in 
Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, our 
version of the \emph{inside}, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing 
$m$ mutations. The latter in Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
 computes the \emph{outside}, in particular,  
the partition function considering only between $[0,i]\cup[j,n-1]$
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing 
 $m$ mutations outside.

\subsubsection{Definitions}
Let be $B:=\left\{\text{A},\text{C},\text{G},\text{U}\right\}$, the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$. Let $R$ be the Boltzmann constant, $T$ the temperature in Kelvin and
  the function $\delta$ such that: 
 $\forall a,a' \in B, \delta_{a,a'}:=\left\{\begin{array}{ll}
															1 & \text{If } a\equiv a'\\
															0 & \text{Otherwise.}
														\end{array}\right.$
 Let us finally denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
\begin{equation}
  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
\end{equation}

\subsubsection{Inside}
\begin{figure}
\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
\caption{Illustration of the principle underlying the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
can be decomposed as a sequence over $[i+1,j]$ preceded by a, possibly mutated, base at $i$ 
(Unpaired case), a sequence over $[i+1,j-1]$ surrounded by some base-pair (Stacking-pair case), 
or as two sequences over $[i+1,k-1]$ and $[k+1,j]$, completed by some base-pair (General base-pairing case). 
In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
and locally instanciated bases.}
\end{figure}

@@

\subsubsection{Inside-Outside}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{forward} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{backward} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

>>>>

\subsection{Probabilistic model}
Our probabilistic model aims at capturing both the stability of the folded RNA and its ability to adopt a predefined 3D conformation.
To that purpose, a Boltzmann weighted distribution is used, based on a pseudo-energy function $\PE{\cdot}$ which includes contributions for both 
the free-energy and the isostericity. 
More precisely, let $\Omega$ be an gapless RNA alignment, $S$ its associated secondary structure, 
then each sequence $s$ has probability proportional to its Boltzmann factor
\begin{align*}
  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT} &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
\end{align*}
where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ 
are the free-energy and isostericity contributions respectively, further described below, and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is the tabulated free-energy of a stacking pair in the Turner model if available, $0$ if $ab=\varnothing$ (no base-pair to stack onto), and using default value $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$). This latter parameter allows to choose whether to completely forbid a sequence inducing invalid base pairs ($\beta = \infty$), or to simply penalize it.
The imprecision introduced by this simplification of the Turner model remains reasonable, since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity contribution}
The concept of isostericity captures the geometric superimposability of two base-pairs, using individual scores given by Stombaugh~\emph{et al}~\cite{Stombaugh2009} based on a probabilistic model. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gapless RNA alignment $\Omega$,  the contribution of the isostericity to the pseudo-energy is given by
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence in comparison with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the alignment sequences.

\subsection{Mutational profile of sequences}

Let $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the 
{SCFG} generating  
all  secondary structures of length  $|s|$. We are interested in  
the probability of a given position being a specific nucleotide,
over all sequences having at most $m$ mutations from $s$, under the SCFG derivation $S$ 
(formally 
$\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define as a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former presented in 
Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, our 
version of the \emph{inside} algorithm, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing 
$m$ mutations. The latter in Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
 computes the \emph{outside}, in particular,  
the partition function considering only between $[0,i]\cup[j,n-1]$
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing 
 $m$ mutations outside.

\subsubsection{Definitions}
Let be $B:=\left\{\text{A},\text{C},\text{G},\text{U}\right\}$, the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$. Let $R$ be the Boltzmann constant, $T$ the temperature in Kelvin and
  the function $\delta$ such that: 
 $\forall a,a' \in B, \delta_{a,a'}:=\left\{\begin{array}{ll}
															1 & \text{If } a\equiv a'\\
															0 & \text{Otherwise.}
														\end{array}\right.$

Finally, we denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
\begin{equation}
  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
\end{equation}

\subsubsection{Inside}
\begin{figure}
\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
\caption{Illustration of the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
can be decomposed as a sequence over $[i+1,j]$ preceded by a, possibly mutated, base at $i$ 
(Unpaired case), a sequence over $[i+1,j-1]$ surrounded by some base-pair (Stacking-pair case), 
or as two sequences over $[i+1,k-1]$ and $[k+1,j]$, completed by some base-pair (General base-pairing case). 
In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
and locally instanciated bases.}
\end{figure}

@@

\subsubsection{Combining Inside and Outside value into base content probabilities}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{forward} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{backward} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$
